Psychophysics of musical elements in the discrete-time representation of sound

### 2013-10-06 article/article.tex
feito ateh intervalos

\section{Notes organization in music}\label{notasMusica} \label{sec:notesMusic}

Consider $ S_j=\left\{  s_j=T_i^j=\{t_i^j\}_{i=0}^{\Lambda_j-1} \right\}_{j=0}^{H-1} $ as the sequence $S_j$ of $H$ musical events $s_j$. In addition, $S_j$, called a 'musical structure', as composed by events $s_j$ which are musical structure themselves, e.g.\ notes. This section is dedicated to techniques that make $S_j$ interesting and enjoyable for audition.

The elements of $S_j$ can be overlapped by mixing them together as in
equation~\ref{eq:mixagem} and Figure~\ref{fig:mixagem}, building intervals
and chords. This scenery reflects the 'vertical thought' in music. On the other hand, the concatenation of $S_j$
as in equation~\ref{eq:concatenacao} and in Figure ~\ref{fig:concatenacao},
builds melodic sequences and rhythms, which are associated with the musical
'horizontal thought' in music. The fundamental frequency $f$ and the beginning moment
(attack) are generally the most important characteristics of the elements in
$S_j$. Such characteristics make possible the creation of music made by pitches (both
harmony and melody) and by the presence of temporal metrics and rhythms.

@@

Here we describe characteristics in music related to harmony and melody.

@@

Doubling the frequency is equivalent to ascending one octave ($f=2f_0$).
The octave division in twelve notes is the canon of classical western music, although its usage has also been observed
outside western tradition, like ceremonial/religious and ethnic context~\cite{Wisnick}.

The factor given by $\varepsilon=2^{\frac{1}{12}}$ defines a semitone. It builds a note
grid along the hearing spectrum. Fixing a frequency $f$, all the possible
fundamental frequencies stay separated by intervals that are multiples of
$\varepsilon$. Twelve semitones (or half steps), equidistant to the human ear, make an
octave. Therefore, if $f=2^{\frac{1}{12}}f_0$, there is a semitone between
$f_0$ and $f$.

This absolute accuracy is common in computational implementations. For the real
music instruments, however, there are deviations of those frequencies in order to make
the harmonics of corresponding notes more compatible. Besides that, the fixed reference
$\varepsilon=2^{\frac{1}{12}}$ characterizes an equally tempered tuning. There
are tunings which respective intervals are ratios of low-order integers, originally based
on observations of physical behaviors. First of these tunings were formalized
around two thousand years before the existence of an equal temperament~\cite{Roederer}.

Two iconic tunings:

\begin{itemize}
    \item The {\bf just intonation} is defined as notes of the diatonic scale
    that are associated with ratios of low-order integers that defines the harmonic
    series. The basic ratios are the same in the Ionian mode (white piano keys
    from \do to \do (see below in subsection~\ref{subsec:escalas}): 1, 9/8, 5/4,
    4/3, 3/2, 5/3, 15/8, 2/1). 
    The intervals are considered as notes in a scale together with the following: the
    semitone 16/15, the 'minor tone' 10/9, and the 'major tone' 9/8. There are
    many ways to perform the division of the 12 notes.

    \item The {\bf Pythagorean tuning} is based on the interval 3/2 (perfect
    fifth). The Ionian mode becomes: 1, 9/8, 81/64, 4/3, 3/2, 27/16, 243/128,
    2/1. The intervals are also considered as notes in a scale. Besides the
    intervals in the mode, it is also used the minor second 256/243, the minor
    third 32/27, the augmented fourth 729/512, the diminished fifth 1024/729,
    the minor sixth 128/81 and the minor seventh 16/9. 
\end{itemize}

In order to attends for micro-tonality\footnote{Micro-tonality is the use of intervals less
than one semitone and have ornamental and structuring functionalities in music. The
division of the octave in $12$ notes is motivated by physical fundamentals but it is also
a \emph{convention} adopted even by western classical music. Other tunings are
present. To cite one example, the traditional Thai music uses a octave
division in seven notes equally spaced ($\varepsilon=2^{\frac{1}{7}}$),
resulting in intervals that have little in common with the intervals in the
diatonic scale~\cite{Wisnick}}. It is possible to use non-integer real values as
pitch sequences or even to maintain the integer values and change the factor $\varepsilon=2^{\frac{1}{12}}$. For example, 
a tuning really near of the harmonic series
itself is proposed with the division of the octave in $53$ notes:
$\varepsilon_2=2^{\frac{1}{53}}$~\cite{microtonalidade}.
In that division the notes are related by means of integers with
$\varepsilon_2$. Note that if $S_i$ is a pitch sequence related by means of
$\varepsilon_1$, a mapping for all the notes related by $\varepsilon_2$ defines
a new sequence $S_i'=\left\{s_i'\right\}=\left\{
s_i \frac{\varepsilon_1}{\varepsilon_2}\right\}$. The music piece \emph{Micro
tom} explores this micro-tonal features and its code is in
Appendix~\ref{ap:micro} and is part of \massa.

>>>>

\section{Organization of notes in music}\label{notasMusica} \label{sec:notesMusic}

Consider $ S_j=\left\{  s_j=T_i^j=\{t_i^j\}_{i=0}^{\Lambda_j-1} \right\}_{j=0}^{H-1} $, the sequence $S_j$ of $H$ musical events $s_j$. Be $S_j$ a `musical structure', composed by events $s_j$ which are musical structures themselves, e.g.\ notes. This section is dedicated to techniques that make $S_j$ interesting and enjoyable for audition.

The elements of $S_j$ can be overlapped by mixing them together, as in
equation~\ref{eq:mixagem} and figure~\ref{fig:mixagem}, for building intervals
and chords. This reflects the `vertical thought' in music. On the other hand, the concatenation of the events
in $S_j$, as in equation~\ref{eq:concatenacao} and in figure ~\ref{fig:concatenacao},
builds melodic sequences and rhythms, which are associated with the musical
`horizontal thought'. The fundamental frequency $f$ and the starting moment
(attack) are generally the most important characteristics of the elements $s_j$ in
$S_j$. This makes possible the creation of music made by pitches (both
harmony and melody) and by the presence of temporal metrics and rhythms.

@@

@@

Doubling the frequency is equivalent to ascending one octave ($f=2f_0$).
The octave division in twelve steps is the canon of classical western music.
Its usage has also been observed
outside western tradition, like ceremonial/religious and ethnic context~\cite{Wisnick}.
In this system, ideally, the factor given by $\varepsilon=2^{\frac{1}{12}}$ defines a semitone.
A note grid along the hearing spectrum in which, given the frequency $f$, all the possible
fundamental frequencies are separated by intervals, multiples of
$\varepsilon$. Twelve semitones (or half steps), equidistant to the human ear, reaches an
octave, equivalent to human year. Therefore, if $f=2^{\frac{1}{12}}f_0$, there is a semitone between
$f_0$ and $f$.

This absolute accuracy is common in computational implementations. Real
music instruments, however, exhibit deviations in order to make compatible
deviations from the harmonics of notes. Besides that, the fixed reference
$\varepsilon=2^{\frac{1}{12}}$ characterizes an equally tempered tuning. There
are tunings in which intervals are ratios of low-order integers, based
on observations of physical phenomena. These tunings were first formalized
around two thousand years before the arrival of the equal temperament~\cite{Roederer}.
). Two emblematic tunings are:

\begin{itemize}
    \item The {\bf just intonation}, defined by association of diatonic scale notes
    with ratios of low-order integers, as found in the harmonic
    series. Considered the ionian mode, the white piano keys
    from C to C, as seen below in subsection~\ref{subsec:escalas}, i.e. intervals 1, 2M, 3M, 4J, 5J, 6M, 7M, 8,
    are considered ratios of frequency: 1, 9/8, 5/4
    4/3, 3/2, 5/3, 15/8, 2/1. 
    The following intervals are also considered: the
    semitone 16/15, the 'minor tone' 10/9, and the 'major tone' 9/8.
    There are many ways to perform the division of the 12 notes in the just intonation.

    \item The {\bf Pythagorean tuning}, based on the interval 3/2 (perfect
    fifth). The Ionian mode ratios becomes: 1, 9/8, 81/64, 4/3, 3/2, 27/16, 243/128,
    2/1. These intervals are also considered to yield notes in a scale. Besides the
    intervals in the mode, also used are the minor second 256/243, the minor
    third 32/27, the augmented fourth 729/512, the diminished fifth 1024/729,
    the minor sixth 128/81 and the minor seventh 16/9. 
\end{itemize}

In order to attend for micro-tonality\footnote{Micro-tonality is the use of intervals less
than one semitone and have ornamental and structuring functionalities in music. The
division of the octave in $12$ notes is motivated by physical fundaments but also
a \emph{convention} adopted too by western classical music. Other tunings are
incident. To point only one example, the traditional Thai music uses an octave
division in seven notes equally spaced ($\varepsilon=2^{\frac{1}{7}}$),
which allows intervals that have little in common with the intervals in the
diatonic scale~\cite{Wisnick}}. It is possible to use non-integer real values as
factors of $\varepsilon=2^{\frac{1}{12}}$ between frequencies, 
or to maintain integer values and change $\varepsilon$. For example, 
a tuning really near of the harmonic series
itself is proposed with the equal division of the octave in $53$ notes:
$\varepsilon_2=2^{\frac{1}{53}}$.~\cite{microtonalidade}
In this division, notes are related by means of integers, factors of
$\varepsilon_2$ relating frequencies by a valid interval.
 Note that if $S_i$ is a pitch sequence related by means of
$\varepsilon_1$, a direct mapping for notes related by $\varepsilon_2$ is 
a new sequence $S_i'=\left\{s_i'\right\}=\left\{
s_i \frac{\varepsilon_1}{\varepsilon_2}\right\}$. The music piece \emph{Micro
tom} explores microtonal features and its code
is part of \massa\ toolbox.
